---
title: "EDA exercises"
output: html_document
---

Read in the fish dataset from the URL.
```{r}
fish <- read.table("http://www.bio.fsu.edu/miller/docs/Tutorials/eg1.txt",head=T)
```

What command would you use to see the variable names in the dataset?

```{r}
#command here
```

The data are derived from a larger study of character variation in a species of fish, sailfin mollies. Specifically, the data consist of selected measurements on females from three populations (identified by the POP variable, populations 1, 2 and 5). Ignore the column IDNO. The actual data recorded include several measurements taken on individual fish: RAYNO, a discrete variable for the number of rays in the dorsal fin; SL, a quantitative variable of the body lengths in mm; FINAREA, a quantitative variable containing measurements of the area of the dorsal fin in sq mm; and TAREA, a quantitative variable of the tail-fin area in sq mm.

Attach the data first
```{r}
attach(fish)
```

##Histogram
Plot histograms of the SL, FINAREA and TAREA variables.
```{r}
#your code here (refer to code on slides if needed)
```

These data come from 3 different populations. Plot histograms of body lengths in population 1 and dorsal fin area from population 5.
```{r}
#your code here
```

Create a customized histogram with detailed axis labels, a main heading and different colors. Type ?hist or help(hist) to learn about the arguments you need to include in the hist() command for these changes.
```{r}
#your code here
```

Create a side-by-side boxplot of SL for each of the 3 populations within the same figure.
```{r}
#your code here
```

Create a scatterplot with TAREA on the x axis and FINAREA on the y axis.
```{r}
#your code here
```

What if we want to add a linear regression line to our plot?
The function `lm` can be used to specify a linear model and `abline` can be used to a line of the slope-intercept form. Note that `lm` takes the form Y ~ X so this is the OPPOSITE of the order you specify when plotting a scatterplot (i.e. x then y).
So to add a simple linear regression line to the plot we type the following line. 
```{r}
plot() #make sure to put your plot code from above here. R can't place the line unless there's a plot already called.
abline(lm(Y~X)) #replace x and y with the appropriate variable names
```

Plot both your boxplot and scatterplot within the same panel. Use your code above and this line below which specifies 1 row of figures and 2 columns.
```{r}
par(mfrow=c(1,2))
#your code here
```


Create parallel box plots (i.e., 1 row & 3 columns; adjust par(mfrow) code from above appropriately) of a given variable for the three different populations in the fish data set. Add the labels "population" to the x axes and "range" to the y axes. Give the boxes in each plot different colors. In addition, give the graph an appropriate title.
```{r}
#your code here
```

Use the par(mfrow=(c(2,3)) format. Plot the graph above, as well as 3 different scatter plots of TAREA vs. FINAREA (for the three different populations) with appropriate axes labels and headings, all on the same graphics panel. Add a linear regression line showing the relationship between the points in each graph and use a different color for the points and lines in each of your three scatter plots.
```{r}
#your code here
```


