---
title: "ggplot2 tutorial"
output: html_document
---

First get the package:
```{r, message=FALSE, warning=FALSE}
#install.package(ggplot2)
library(ggplot2)
```

Let's say we wanted to plot some two-variable data, changing color and shape by the sub-category of data. Here's a data set on plants:
```{r}
data.df <- data.frame(Plant = c("Plant1", "Plant1", "Plant1", "Plant2", "Plant2", 
    "Plant2"), Type = c(1, 2, 3, 1, 2, 3), Axis1 = c(0.2, -0.4, 0.8, -0.2, -0.7, 
    0.1), Axis2 = c(0.5, 0.3, -0.1, -0.3, -0.1, -0.8))
data.df
```

Using normal R graphics, we might do this:
```{r}
color_foo <- colorRampPalette(c("lightblue", "darkblue"))
colors <- color_foo(3)

plot(range(data.df[, 3]), range(data.df[, 4]), xlab = "Axis 1", ylab = "Axis 2", 
    type = "n")
points(data.df$Axis1, data.df$Axis2, pch = c(1, 2)[data.df$Plant], col = colors[data.df$Type])
legend("topright", legend = c("Plant1", "Plant2"), pch = 1:2)
legend("bottomright", legend = c("Type1", "Type2", "Type3"), pch = 20, col = colors)
```

With `ggplot`, you just do this:
```{r}
ggplot(data.df, aes(x = Axis1, y = Axis2, shape = Plant, color = Type)) + geom_point(size = 5)
```

`ggplot` is best used on data in the data frame form. Let's look at a data set already in R (msleep). This is about the sleep habits of different animal species.
```{r}
data(msleep)
head(msleep)
```

Let's start with a basic plot. Let's create a scatterplot of body weight against total hours sleep.
```{r}
a <- ggplot(data = msleep, aes(x = bodywt, y = sleep_total))
a <- a + geom_point()
a <- a + xlab("Body Weight") + ylab("Total Hours Sleep") + ggtitle("Some Sleep Data")
a
```

So here's what we just did. The ggplot() command creates a plot object. In it we assigned a data set. aes() creates an aesthetic: a mapping of variables to various parts of the plot. We then add components to the plot. geom_point() adds a layer of points, using the base aesthetic mapping. The third line adds labels. Typing the variable name `a` displays the plot. Or, you can use the command ggsave() to save the plot as a file, as in the code below. The plot will automatically be saved in the working directory once the line is run.
```{r}
ggsave("sleepplot.png", plot = a, width = 4, height = 4)
```

Now, one of the great things we can do with ggplot is slice the data different ways. For instance, we can plot another variable against color:
```{r}
a <- ggplot(data = msleep, aes(x = bodywt, y = sleep_total, col = vore))
a <- a + geom_point()
a <- a + xlab("Body Weight") + ylab("Total Hours Sleep") + ggtitle("Some Sleep Data")
a
```

What changed in the code to get this new variable `vore` plotted with colors?

We can transform variables directly in the ggplot call, as well:
```{r, message=FALSE, warning=FALSE}
a <- ggplot(data = msleep, aes(x = log(bodywt), y = sleep_rem/sleep_total, col = vore))
a <- a + geom_point()
a <- a + xlab("Log Body Weight") + ylab("Fraction of Sleep that is REM") + ggtitle("Some Sleep Data") + 
    scale_color_discrete(name = "Trophic Level")
a
```

Within the geom calls, we can change plotting options:
```{r, warning=FALSE, message=FALSE}
a <- ggplot(data = msleep, aes(x = log(bodywt), y = sleep_rem/sleep_total, col = vore))
a <- a + geom_point(size = 5) #we are adding an argument here adjust the size of the points used
a <- a + xlab("Log Body Weight") + ylab("Fraction of Sleep that is REM") + ggtitle("Some Sleep Data") + 
    scale_color_discrete(name = "Trophic Level")
a
```

Let's use a different data set to do line plots:
```{r}
data(economics)
head(economics)
```

Line plots:
```{r}
a <- ggplot(data = economics, aes(x = date, y = unemploy))
a <- a + geom_line()
a
```

We can add statistical transformations to this series, for instance:
```{r}
a <- ggplot(data = economics, aes(x = date, y = unemploy))
a <- a + geom_line()
a <- a + geom_smooth()
a ## geom_smooth: method="auto" and size of largest group is <1000, so using loess. Use 'method = x' to change the smoothing method.
```

##In-class Exercises
Using msleep data and `ggplot2` package:

1. Make a scatterplot of `sleep_total` vs. `sleep_rem` (REM on x axis) with appropriate axes labels and title.
```{r}
#your code here
```

2. Make point size proportional to log(bodywt).
```{r}
#your code here
```

3. Add a OLS regression line to the plot from #2.
```{r}
#your code here
```

4. Color-code the points from plot #2 according to vore (trohpic level).
Adjust the legend labels so they are easily understood (i.e. change from vore)
Does the scaling of REM & total sleep differ with diet (provide answer below your graph)?
```{r}
#your code here
```

5. Create boxplots of `sleep_total` by `vore`. Adjust axes labels so they are easily understood, including the tick mark labels (i.e. spell out carnivore, etc.). Hint: type `?scale_x_discrete` to start with the x-axis.
```{r}
#your code here
```

6. Overlay points on top of the boxplots in a color of your choice.
```{r}
#your code here
```

