---
title: "Binomial Distribution"
output: html_document
---

##Using dbinom, pbinom, rbinom, qbinom
There are four functions that can be used to generate the values associated with the binomial distribution. You can get a full list of them and their options using the help command:
```{r}
#help(Binomial)
```

This will open a help webpage describing the command.

These commands work just like the commands for the normal distribution. The binomial distribution requires two extra parameters, the number of trials and the probability of success for a single trial. The commands follow the same kind of naming convention, and the names of the commands are dbinom, pbinom, qbinom, and rbinom.

A few examples are given below to show how to use the different commands. First we have the distribution function, dbinom:
```{r}
x <- seq(0,50,by=1)
y <- dbinom(x,50,0.2)
plot(x,y)
y <- dbinom(x,50,0.6)
plot(x,y)
y <- dbinom(x,100,0.9)
plot(x,y)
```

How does changing p (probability) alter the distributions above?

Next we have the cumulative probability distribution function. `pbinom` takes 3 arguments: number of successes, number of trials, probability of success. 

```{r}
pbinom(25,50,0.5)

pbinom(50,50,0.9)

pbinom(25, 100, 0.5)

```

How does changing each of the 3 numbers alter the results?

Finally random numbers can be generated according to the binomial distribution, where you specify the number of observations to generate, the number of trials, and the probability of success on each trial.

```{r}
rbinom(5,100,.2)
rbinom(5,100,.7)

```

##In-class Binomial probability exercises
1. You roll 12 dice simultaneously and let X denote the number of 6's that appear. We want to find the probability of getting 7, 8 or 9 6's. If we let S=[get a 6 on one roll], then P(S)=1/6 and the rolls consitute individual Bernoulli trials. Thus, X~binom(size=12,prob=1/6). We need to find P(7<= X >= 9). Include your code below.

```{r}
#your code here
```

2. We can use R to simulate the process of drawing random samples from a binomial distribution. For example, pretend you drew 6 cards from a randomly shuffled deck and counted cards with a value of 1-7 as a successful 'mating' of Drosophila in your study system. You repeated this 20 times. To simulate this in R, you would use the `rbinom` function. Indicate your code below.

Note: There are 52 cards in a deck and 4 suits of each number.
```{r}
#your code here
```

3. Plot a histogram of the distribution from the step above. Adjust the breaks argument such that there are an appropriate number of breaks in your histogram.

What does the histogram illustrate and how would this change if you counted cards with a value of 1-3 as a successful 'mating' instead?

```{r}
#your code here
```

4. Set up your graphing window to hold 4 plots using the par(mfrow()) command. Then, plot (using lines instead of points) the binomial densities for the following values of n and p: (n=20,p=0.5); (20, 0.25); (20, 0.75); (20, 9/10). Where is the peak density  (the mode) located in each case? What is the relationship between the mode and the values of n and p?

```{r}
#your code here
```

