---
title: "Probability Exercises"
output: html_document
---

##Using pnorm command: normal distributions and densities
`pnorm` has the format: # of interest, mean, SD.
By default, `pnorm` calculates less than the # of interest. To do greater than, include the argument lower.tail=FALSE

## 1. Probability female less than 60 inches?
Mean height of females is 64 in; SD is 3 in.
```{r}
pnorm(60,64,3,lower.tail=TRUE)
```

## 2. Probability male at least 72 inches?
Mean height of males is 70 in; SD is 3 in.
```{r}
#your code here
```

## 3. Height such that 2% of females shorter than x?
The command `qnorm` let's you work with a probability, not a quantile like with `pnorm`.
```{r}
qnorm(0.02,64,3,lower.tail=TRUE)
```

## 4. Proportion of males between 68 and 71 inches?
Hint: you can add/subtract pnorm commands!
```{r}
#your code here
```

## 5. Distribution for males
```{r, echo=FALSE}
x<-seq(50,80,length=1000) #generates a sequence of 1000 #s btw 50 & 80
y<-dnorm(x,mean=70,sd=3) #generates the density distribution of x with mean of 70 and sd of 3 (males)
plot(x,y,type="l",lwd=1) #plots the density

```

## 6. Adding distribution for females
How would you adjust the code from #5 for females? Do so here and graph the curve on the same plot as the male curve above. 
Don't forget to add a legend (use legend command) and adjust the weight (lwd argument) of the plotted line so that it's heavier for females (so you can distinguish between the 2 curves easily).
```{r,echo=FALSE}

```

