# A Minimal Example for Markdown

This is a minimal example of using **knitr** to produce an _HTML_ page from _Markdown_.

## R code chunks

```{r setup}
#Install rmarkdown and knitr packages first
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5) #set global chunk options: images will be 7x5 inches
```

Now we write some code chunks in this markdown file:

```{r}
x <- 1+1 # a simple calculator

rnorm(5) # boring random numbers
```

We can also produce plots:

```{r graphics}
par(mar = c(4, 4, .1, .1))
with(mtcars, {
  plot(mpg~hp, pch=20, col='darkgray')
  lines(lowess(hp, mpg))
})
```

## Inline code

Inline R code is also supported, e.g. the value of `x` is `r x`, and 2 &times; &pi;
= `r 2*pi`.

## Math

LaTeX math for fancy output: $f(\alpha, \beta) \propto x^{\alpha-1}(1-x)^{\beta-1}$.

## Misc

You can indent code chunks with numbers so they can nest within other environments such as lists.

1. the area of a circle with radius x
    ```{r}
    pi * x^2
    ```
2. Yay, it works!

 
## Conclusion

Markdown is super easy to write. Go to **knitr** [homepage](http://yihui.name/knitr) for details.
 